---
title: "Downstream analysis"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

CLIMB's output is effectively the posterior MCMC samples of the parameters of the normal mixture. These samples can be used for clustering/classification, visualizations, and statistical inference. Here, we show implementations of several of the analyses used in the CLIMB manuscript.

First, let's load in the MCMC we briefly ran on simulated data in the [running the MCMC](running_mcmc.html) section.

```{r}
chain <- readRDS("output/chain.rds")
```

## Visualizing convergence

The first thing we should do make some trace plots for some of the parameters being estimated. This will help us decide when the chain has entered its stationary distribution. Here, we make plots of a mean, variance, covariance, and mixing weight.

```{r, fig.height = 4, fig.width=16, echo = 3:13}
par(mfrow = c(1,4))

# The mean in the first dimension of the third cluster
mu31 <- chain$mu_chains[[3]][,1]

# The variance of the first dimension of the second cluster
sigma21 <- chain$Sigma_chains[[2]][1,1,]

# The covariance between dimensions 2 and 3 in the fourth cluster
sigma423 <- chain$Sigma_chains[[4]][2,3,]

# The mixing weight of the seventh cluster
pi7 <- chain$prop_chain[,7]

plot(mu31, type = "l", xlab = "MCMC iteration", ylab = expression(hat(mu)[31]))
plot(sigma21, type = "l", xlab = "MCMC iteration", ylab = expression(hat(Sigma)[21]))
plot(sigma423, type = "l", xlab = "MCMC iteration", ylab = expression(hat(Sigma)[423]))
plot(pi7, type = "l", xlab = "MCMC iteration", ylab = expression(hat(pi)[7]))

```

Looking at a few of these trace plots reveals that the chain achieves stationary around 200 iterations in. Therefore, we discard the first 200 iterations as burn-in, and do our analyses on the remaining 800.
```{r}
burnin <- 1:200
```

## Clustering dimensions and observations

CLIMB can provide a **bi-clustering** of analyzed data. Row clusters come from the estimated latent classes, of course, though these classes can even be grouped by their similarity.

CLIMB uses a distance based on class covariances to cluster the dimensions (columns). The function `compute_distances_between_conditions` will compute these pairwise distances between conditions based on model fit. This was used in the CLIMB manuscript to determine the relationship among hematopoietic cell populations based on CTCF binding patterns. 

CLIMB's estimated classes naturally cluster the rows. However, the function `compute_distances_between_clusters` will compute distances between estimated clusters as well. This distance is based on a symmetrified theoretical Kullback-Liebler divergence. 

```{r, echo = 3:6, fig.height=3.5, fig.width=7.5}
library(CLIMB)
library(purrr)
par(mfrow = c(1,2))

# The figure isn't as exciting with simulated data
col_distmat <- compute_distances_between_conditions(chain, burnin)
hc_col <- hclust(as.dist(col_distmat), method = "complete")
plot(hc_col, xlab = "", ylab= "", axes = FALSE, sub = "", main = "Row cluster dendrogram")

row_distmat <- compute_distances_between_clusters(chain, burnin)
hc_row <- hclust(as.dist(row_distmat), method = "complete")
plot(hc_row, xlab = "", ylab= "", axes = FALSE, sub = "", main = "Column cluster dendrogram")
```

These figures aren't very thrilling on their own, but we can use them to make some nice heatmaps. CLIMB provides the function `get_row_reordering` to facilitate plotting the bi-clustering:

```{r, echo = -1, fig.width = 2.5, fig.height = 6}
library(ggplot2)

# Load back in the data
data("sim")
dat <- sim$data

# Get a row reordering for plotting
row_reordering <- get_row_reordering(row_clustering = hc_row, chain = chain, burnin = burnin)

# Melt (for ggplot2)
molten1 <-
  reshape2::melt(data = dplyr::mutate(dat, "row" = row_reordering),
  id.vars = "row")

# Rename the factor levels to match the dimensions
levels(molten1$variable) <- paste(1:3)[hc_col$order]

# Plot the bi-clustering heatmap, like in the CLIMB manuscript
p1 <- ggplot(data = molten1,
            aes(x = variable, y = row, fill = value)) +
    geom_raster() +
    cowplot::theme_cowplot() +
    theme(legend.position = "none", axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
    xlab("") + ylab("") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = .5)) +
    ggtitle("Bi-clustering\nheatmap") +
    scale_fill_gradientn(colours=RColorBrewer::brewer.pal(9,"Greens"))

print(p1)
```


## Merging groups

CLIMB allows you to merge similar groups for the sake of visualization, for examble on the Genome Browser. In the CLIMB manuscript, we merged clusters of CTCF binding patterns to color peaks more in a more parsimonious set of 5 groups on the genome browser. We show how to do this for an arbitrary number of groups with CLIMB's `merge_classes` function.

The output of a `merge_classes` call provides the means and mixing weights of the new parent groups. These means and weights are aggregated from each groups constituent classes. `merged_z` gives the **new group assignments** for each observations. The `clustering` obect is simply a call to `stats::hclust()`.

```{r}
# Merge into 5 parent groups
merged <- merge_classes(n_groups = 5, chain = chain, burnin = burnin)

# Check the structure of the merging
str(merged)
```

We can visualize the salient features of the new groups with a simple visualization of the merged group means:

```{r, fig.width = 5, fig.height = 2.5}
# For the plot color palette
library(RColorBrewer)

# Melt for plotting
molten2 <- reshape2::melt(merged$merged_mu) %>%
  `colnames<-` (c("group", "condition", "mu"))

p2 <- ggplot(data = molten2, aes(group, condition, fill = mu)) +
  geom_tile(color = "black") +
  scale_fill_distiller(palette = "PiYG", direction = 1) +
  coord_fixed() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, vjust = 0.5, hjust = 1, size = 18),
          axis.text.y = element_text(size = 20),
          axis.title.x = element_text(size = 18),
          axis.ticks.y = element_blank(),
          axis.ticks.x = element_blank(),
          legend.title = element_text(size = 12))

print(p2)
```

## Testing for consistency



### To be written...


