---
title: "Pairwise fitting"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

### <i><b><span style="color:salmon">---Special considerations: this portion is highly parallelizable---</span></b></i>

Here, we describe how to execute the first step of CLIMB: pairwise fitting (a composite likelihood method).

First, load the package and the simulated dataset. This **toy** dataset has $n=1500$ observations across $D=3$ conditions (*that is,* dimensions). Thus, we need to fit $\binom{D}{2}=3$ pairwise models.

```{r, echo = FALSE}
knitr::opts_chunk$set(autodep = TRUE)
```

```{r}
# load that package
library(LIEB)

# load the toy data
data("sim")
```

The fitting of each pairwise model can be done in parallel, which saves a lot of computing time when the dimension is larger. This can be done simply (in parallel, or linearly) with the function `get_pairwise_fits()`. Note that the input data should be $z$-scores (or data arising from some other scoring mechanism, transformed appropriately to $z$-scores).

`get_pairwise_fits()` runs the pairwise analysis at the default settings used in the CLIMB manuscript. The user can select a few settings with this functions:

1. `nlambda`: how many tuning parameters to try (defaults to 10)

2. `parallel`: logical indicating whether or not to do the analysis in parallel

3. `ncores`: if in parallel, how many cores to use (defaults to 10)

4. `bound`: is there a lower bound on the estimated non-null mean? (defaults to zero, and must be non-negative)

With all of this in place, one can obtain the pairwise fits as follows:
```{r, cache = TRUE, message = FALSE, warning = FALSE}
fits <- get_pairwise_fits(z = sim$data, parallel = FALSE)
```

Calling `names(fits)` tells us which pair of dimensions each fit belongs to. 
```{r}
names(fits)
```

Each fit contains additional information, including the length-2 association patterns estimated to be in the given pairwise fit, the posterior probability of each observation belonging to each of these classes, and their corresponding estimated means and covariances.

Finally, save this output, as it is necessary for many parts of the downstream analyses, before moving on to [the next step](candidate_latent_classes.html).

```{r, eval = FALSE}
save(fits, file = "pwfits.Rdata")
```